---
title: "Selecting data columns"
output: learnr::tutorial
runtime: shiny_prerendered
description: Short demo converting CSIRO Data School course notes
---

```{r setup, include=FALSE}
library(learnr)
library(tidyverse)
library(gapminder)
gapminder <- gapminder::gapminder
knitr::opts_chunk$set(echo = TRUE)
```

## Selecting columns from a data frame

The first task we will cover is extracting specific columns from a dataset to work with. This 
may be needed if your data is very, very wide and you are only interested in a few columns. The 
gapminder data doesn't exactly meet that description, but we will continue using it to understand 
how the functions work.

```{r load-gapminder, eval=F}
gapminder <- read_csv("data/gapminder.csv")
```

### The `select()` function

You can specify columns to keep from a data frame using `select()`. This function, as well as many 
others we will be learning today are found in the `dplyr` package of the tidyverse. At it's simplest,
you provide `select()` with a data frame and the column names you would like to keep.

```{r select-by-name}
#Select the year, country and pop columns
select(gapminder, year, country, pop)
```

Here you can see that the `select()` produces a new data frame with just the columns you provided. 
It also keeps them in the order you specified.

You can also select columns by index position:

```{r select-by-pos}
#Select the first, third and fourth columns
select(gapminder, 1, 3, 4)
```

or you can select columns to *exclude* by using negation:

```{r select-by-negation}
#Remove lifeExp and pop columns
select(gapminder, -lifeExp, -pop)
```

If the columns you want are all together in the original data frame, you can select them as a range

```{r select-range}
#Select all columns between year and lifeExp
select(gapminder, year:lifeExp)

#Exclude all columns between year and lifeExp
select(gapminder, -(year:lifeExp))
```

### Save your output
In the examples above, the output of the `select()` function is being printed to the screen to 
demonstrate how it works. Normally, you will want to actually *use* the selected data later on, so
don't forget to save the output as a variable.
```{r select-save-variable}
# Just prints to screen, can't use data
select(gapminder, year, country, pop)

# Saves output to variable; can be accessed later
just_population <- select(gapminder, year, country, pop)

just_population
```


### Hands on with `select()`
Use three different methods to select just the `country`, `year`, `pop`, and `gdpPercap` columns 
from the `gapminder` dataframe.
```{r chal-1, exercise=T, exercise.cap = "Using select()"}
select(gapminder, )

select(gapminder, )

select(gapminder, )
```

<div id="chal-1-hint">
You can specify columns to `select()` by **name**, **position**, **negation**, or **range**.
</div>

```{r chal-1-solution}
#By name
select(gapminder, country, year, pop, gdpPercap)

#By position
select(gapminder, 1, 3, 5, 6)

#By exclusion
select(gapminder, -continent, -lifeExp)
```


## Select helper functions

There are a number of helper functions that can be used to select the correct columns. Some commonly
used ones include `starts_with()`, `ends_with()` and `contains()`, but you can see a full list by 
looking at the help file (`?tidyselect::select_helpers`).

These helpers do exactly what you would expect from their names, but to see them in action:

```{r select-helpers}
#All columns that start with "co"
select(gapminder, starts_with("co"))

#All columns that contain the letter "e"
select(gapminder, contains("e"))
```

### Using `select()` helpers
Select all columns from `gapminder` that end with the letter "p"
```{r chal-2, exercise = T, exercise.cap = "Using helpers"}
select(gapminder, )
```

```{r chal-2-solution}
select(gapminder, ends_with("p"))
```


### Unexpected output
Observe the output of the following code. Does it behave the way you expected? What additional argument do you need to provide to `contains()`
to make it return only the `gdpPercap` column?
~~~~

~~~~
{: .language-r}
```{r chal-3, exercise = T, exercise.eval = T}
select(gapminder, contains("P"))
```

<div id="chal-3-hint">
The `contains()` function is actually provided by the `tidyselect` package, try 
`?tidyselect::contains` if you are having trouble finding the help page for it.
</div>

```{r chal-3-solution}
select(gapminder, contains("P", ignore.case = F))
```

## Renaming columns
### `rename()`

`select()` *can* be used to rename columns while you are selecting them using the form 
`new_name = old_name`:

```{r select-rename}
select(gapminder, country_name = country, population = pop)
```

but since it drops all the other columns, it is might not be the outcome you were looking for. 
Instead, `rename()` can rename columns while retaining the rest of the data frame.

```{r rename-example}
rename(gapminder, country_name = country, population = pop)
```

This will often be useful to convert column names from your imported data into the 
[consistent naming format]({{ page.root }}{% link _episodes/02-Using-R.md %}#naming-things) you
have decided on.

### Renaming things
Rename the `lifeExp` and `gdpPercap` columns to the `snake_case` format
```{r chal-4, exercise = T}
rename(gapminder, )
```

```{r chal-4-solution}
rename(gapminder, life_exp = lifeExp, gdp_per_cap = gdpPercap)
```

